---
title: "Exploratory Cold Shock Markers"
output: html_document
author: "Kourosh Zarringhalam & Yasaman Rezvani"
header-includes: \usepackage{xcolor}

date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```

```{r loadLibs, echo=FALSE, message=FALSE, warning=FALSE}
library(Seurat)
library(openxlsx)
library(ggplot2)
library(ggfortify)
library(jcolors)
require(gridExtra)
library(grid)
library(matrixStats)
library(tidyverse)
library(tidytext)
library(RColorBrewer)
library(cowplot)
library(gam)
library(princurve)
library(parallel)
library(sme)
library(plotly)
library(patchwork)
library(ggVennDiagram)
library(UpSetR)
library(ComplexHeatmap)
library(circlize)


source('./loadRDS_coldShock.R')
source('./util_funcs.R')



num.cores <- detectCores(all.tests = FALSE, logical = TRUE)
```

### Integrated Data Processing
1. Data was integrated and processed to identify initial clusters. 
    * Resolution was set to get roughly 4 clusters.
    * Matched clusters accross datasets were retained
    
1. Subsequent data analysis was performed on **individually normalized** data-sets.



```{r Inital_Integration, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=16, fig.height=10}
p1 <- DimPlot(all.samples.integrated, reduction = "pca", 
              split.by = 'spp',
              pt.size = 1,
              #shape.by='spp',
              label = TRUE, label.size = 5) + NoLegend() + 
  theme(panel.spacing = unit(0.5, "lines")) + 
  theme(axis.text.x = element_text(face="bold", size=12, angle=0)) +
  theme(axis.text.y = element_text(face="bold", size=12, angle=0)) +
  theme(
    axis.title.x = element_text(size=14, face="bold"),
    axis.title.y = element_text(size=14, face="bold")
  )

wrap_plots(p1)

```

### Global Marker analysis 

1. Differential Expression Analysis was performed independent of cell cycle phase using WT as base line.


```{r Global_Markers, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
global.shock.markers.sig <- global.shock.markers %>% 
  dplyr::filter(p_val_adj < 0.05 & abs(avg_log2FC) > log2(1.5)) %>% arrange(desc(abs(avg_log2FC)))

global.shock.markers.overlap <- global.shock.markers.sig %>% group_by(GeneID) %>% 
  summarise(num.contrasts = length(query)) %>% arrange(desc(num.contrasts))

global.shock.markers.sig <- left_join(global.shock.markers.sig, global.shock.markers.overlap, by = 'GeneID')
global.shock.markers.sig <- left_join(global.shock.markers.sig, prod.desc, by = 'GeneID')

tmp <- global.shock.markers.sig %>% dplyr::filter(avg_log2FC > 0) %>% group_by(ref, query) %>%
  summarise(genes = list(GeneID))

venn.list <- tmp$genes
names(venn.list) <- paste(tmp$ref, "_vs_", tmp$query, sep = '')

sets <- fromList(venn.list)

upset(sets, nsets = 10, number.angles = 0, point.size = 3.5, line.size = 2)
grid.text(
  "upregulated",
  x = 0.90,
  y = 0.05,
  gp = gpar(
    fontsize = 10,
    fontface = 3
  )
)

ggVennDiagram(venn.list)
```

```{r Global_Markers2, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
tmp <- global.shock.markers.sig %>% dplyr::filter(avg_log2FC < 0) %>% group_by(ref, query) %>%
  summarise(genes = list(GeneID))

venn.list <- tmp$genes
names(venn.list) <- paste(tmp$ref, "_vs_", tmp$query, sep = '')


sets <- fromList(venn.list)

upset(sets, nsets = 10, number.angles = 0, point.size = 3.5, line.size = 2)
grid.text(
  "upregulated",
  x = 0.90,
  y = 0.05,
  gp = gpar(
    fontsize = 10,
    fontface = 3
  )
)
ggVennDiagram(venn.list)
```

### Expression of top genes 

1. Up regulation is mild but steady.

1. Proportion of cells expressing is not changing much.

```{r global_expression_up, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

global.shock.markers.top <- global.shock.markers.sig  %>% ungroup() %>% dplyr::filter(avg_log2FC > 0.6) %>%  
  arrange(desc(num.contrasts), avg_log2FC) %>% transmute(gene = gene, num.contrasts = num.contrasts) %>%
  distinct() %>% head(n=3)
  


DefaultAssay(all.samples.integrated) <- 'RNA'
Idents(all.samples.integrated) <- 'spp' 

all.samples.integrated@meta.data$spp <- factor(all.samples.integrated@meta.data$spp , 
                                               levels = c("BDiv0hrN", "BDiv4hrN", "BDiv12hrN", "BDiv36hrN", 
                                                          "BDiv7dN", "BDiv36hrY", "BDiv7dY"))
all.samples.integrated.sub <- subset(all.samples.integrated, idents = c("BDiv0hrN", "BDiv4hrN", "BDiv12hrN", "BDiv36hrN", "BDiv7dN"))
p2 <- FeaturePlot(object = all.samples.integrated.sub, 
                  shape.by = 'spp',
                  split.by = 'spp',
                  label = T, pt.size = 0.6, label.size = 4, 
                  features = global.shock.markers.top$gene,
                  cols = c("lightgrey", "red"), reduction = "pca") 

plot(p2)
```

```{r global_expression_up2, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
p3 <- DotPlot(all.samples.integrated, features = global.shock.markers.top$gene, 
              cols = c("blue", "red", "green", "yellow"),
              dot.scale = 8) + RotatedAxis()

plot(p3)
```

```{r global_expression_up3, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
VlnPlot(object = all.samples.integrated, features = global.shock.markers.top$gene)
```

1. Down regulation is stranger
1. Proportion of cells expressing declines more.

```{r global_expression_down, eval = F, echo = F, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}

global.shock.markers.top <- global.shock.markers.sig  %>% ungroup() %>% dplyr::filter(avg_log2FC < -0.6) %>%  
  arrange(desc(num.contrasts), avg_log2FC) %>% transmute(gene = gene, num.contrasts = num.contrasts) %>%
  distinct() %>% head(3)

DefaultAssay(all.samples.integrated) <- 'RNA'
Idents(all.samples.integrated) <- 'spp' 
p2 <- FeaturePlot(object = all.samples.integrated.sub, 
                  shape.by = 'spp',
                  split.by = 'spp',
                  label = T, pt.size = 0.6, label.size = 3, 
                  features = global.shock.markers.top$gene,
                  cols = c("lightgrey", "red"), reduction = "pca") 

plot(p2)
```

```{r global_expression_down2, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
p3 <- DotPlot(all.samples.integrated, features = global.shock.markers.top$gene, 
              cols = c("blue", "red", "green", "yellow"),
              dot.scale = 8) + RotatedAxis()

plot(p3)
```

```{r global_expression_down3, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
VlnPlot(object = all.samples.integrated, features = global.shock.markers.top$gene)
```


### Clustering DEGs

1. Hierarchical clustering was performed and expression heatmaps were generated

```{r Global_Heatmap, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=6, fig.height=6}
Idents(all.samples.integrated) <- 'spp'
DefaultAssay(all.samples.integrated) <- 'RNA'
global.avg.expr <- as.data.frame(log1p(AverageExpression(all.samples.integrated, verbose = FALSE)$RNA))
global.avg.expr$gene <- rownames(global.avg.expr)
global.avg.expr$GeneID <- gsub('-', '_', global.avg.expr$gene)

global.up.reg <- global.shock.markers.sig$GeneID[global.shock.markers.sig$avg_log2FC > 0]
global.down.reg <- global.shock.markers.sig$GeneID[global.shock.markers.sig$avg_log2FC < 0]

global.avg.expr.up <- global.avg.expr[global.avg.expr$GeneID %in% global.up.reg, ]
global.avg.expr.down <- global.avg.expr[global.avg.expr$GeneID %in% global.down.reg, ]


global.mat_up <- global.avg.expr.up[,1:(ncol(global.avg.expr.up) - 2)] %>% dplyr::select(!contains('Y')) %>% as.matrix()
global.mat_down <- global.avg.expr.down[,1:(ncol(global.avg.expr.down) - 2)] %>% dplyr::select(!contains('Y')) %>% as.matrix()


L1 <- makeGlobalHeatMap(global.mat_up, prod.desc, 7)
L2 <- makeGlobalHeatMap(global.mat_down, prod.desc, 7)

draw(L1$ht_list)
draw(L2$ht_list)
```



### Matched Marker analysis

1. Integrated data was used to identify matched clusters.

1. In each data sets, markers of clusters were identified

1. Cluster markers were pooled across datasets

1. For each cluster differential expression was performed on pooled marker of the cluster using WT as ctr

```{r matched_clusters, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=16, fig.height=10}
Idents(all.samples.integrated) <- "phase.cond"
DefaultAssay(all.samples.integrated) <- 'RNA'
p1 <- DimPlot(all.samples.integrated, reduction = "pca", 
              split.by = 'spp',
              pt.size = 1,
              #shape.by='spp',
              label = TRUE, label.size = 6) + NoLegend() + 
  theme(panel.spacing = unit(0.5, "lines")) + 
  theme(axis.text.x = element_text(face="bold", size=12, angle=0)) +
  theme(axis.text.y = element_text(face="bold", size=12, angle=0)) +
  theme(
    axis.title.x = element_text(size=14, face="bold"),
    axis.title.y = element_text(size=14, face="bold")
  )

plot(p1)

```


```{r matched_clusters_DEGs, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
matched.DEGs.sig <- matched.DEGs %>% dplyr::filter(p_val_adj < 0.05 & abs(avg_log2FC) > log2(1.5)) %>% arrange(desc(abs(avg_log2FC)))

matched.DEGs.overlap <- matched.DEGs.sig %>% group_by(GeneID) %>% 
  summarise(num.contrasts = length(query)) %>% arrange(desc(num.contrasts))

matched.DEGs.sig <- left_join(matched.DEGs.sig, matched.DEGs.overlap, by = 'GeneID')
matched.DEGs.sig <- left_join(matched.DEGs.sig, prod.desc, by = 'GeneID')

tmp <- matched.DEGs.sig  %>% dplyr::filter(avg_log2FC > 0) %>% group_by(ref, query) %>%
  summarise(genes = list(GeneID))

venn.list <- tmp$genes
names(venn.list) <- paste(tmp$ref, "_vs_", tmp$query, sep = '')

sets <- fromList(venn.list)

upset(sets, nsets = 10, number.angles = 0, point.size = 3.5, line.size = 2)
grid.text(
  "upregulated",
  x = 0.90,
  y = 0.05,
  gp = gpar(
    fontsize = 10,
    fontface = 3
  )
)
```

```{r matched_clusters_DEGs2, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
tmp <- matched.DEGs.sig  %>% dplyr::filter(avg_log2FC < 0) %>% group_by(ref, query) %>%
  summarise(genes = list(GeneID))

venn.list <- tmp$genes
names(venn.list) <- paste(tmp$ref, "_vs_", tmp$query, sep = '')

sets <- fromList(venn.list)

upset(sets, nsets = 10, number.angles = 0, point.size = 3.5, line.size = 2)
grid.text(
  "upregulated",
  x = 0.90,
  y = 0.05,
  gp = gpar(
    fontsize = 10,
    fontface = 3
  )
)

```


### Expression of top genes 

1. up regulated
```{r matched_expression_up, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=16, fig.height=16}

matched.DEGs.top <- matched.DEGs.sig %>% dplyr::filter(avg_log2FC > 0) %>% group_by(cluster) %>% top_n(1, abs(avg_log2FC))
  

sample.ord <- as.numeric(
  gsub('hrY', '', gsub('hrN', '', gsub('BDiv', '', gsub('_.*', '', gsub('7d', '168hr', unique(all.samples.integrated$phase.cond))))))
  )
clust.ord <- as.numeric(gsub('.*_', '', unique(all.samples.integrated$phase.cond)))

re.ord <- rep(c(0:(length(unique(sample.ord)) - 1)), rle(sample.ord[order(sample.ord)])$lengths)
re.ord <- re.ord[order(sample.ord)]
#col.orders <- clust.ord + re.ord * length(unique(clust.ord)) + 1
phase.cond.orders <- order(re.ord +  clust.ord * length(unique(re.ord)) + 1)

all.samples.integrated$phase.cond <- factor(all.samples.integrated$phase.cond, 
                                            levels = unique(all.samples.integrated$phase.cond)[phase.cond.orders])


DefaultAssay(all.samples.integrated) <- 'RNA'
Idents(all.samples.integrated) <- 'spp' 

all.samples.integrated.sub <- subset(all.samples.integrated, idents = c("BDiv0hrN", "BDiv4hrN", "BDiv12hrN", "BDiv36hrN", "BDiv7dN"))

Idents(all.samples.integrated) <- 'phase.cond' 
Idents(all.samples.integrated.sub) <- 'phase.cond' 

p2 <- FeaturePlot(object = all.samples.integrated.sub, 
                  shape.by = 'spp',
                  split.by = 'spp',
                  label = T, pt.size = 0.6, label.size = 3, 
                  features = unique(matched.DEGs.top$gene),
                  cols = c("lightgrey", "red"), reduction = "pca") 

plot(p2)
```

```{r matched_expression_up2, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=8, fig.height=8}
p3 <- DotPlot(all.samples.integrated.sub, features = unique(matched.DEGs.top$gene), 
              cols = c("blue", "red", "green", "yellow"),
              dot.scale = 8) + RotatedAxis()

plot(p3)
```

```{r matched_expression_up3, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
VlnPlot(object = all.samples.integrated.sub, features = unique(matched.DEGs.top$gene))
```

1. down regulated
```{r matched_expression_down, eval = T, echo = F, message=FALSE, warning=FALSE,fig.width=16, fig.height=16}

matched.DEGs.top <- matched.DEGs.sig %>% dplyr::filter(avg_log2FC < 0) %>% group_by(cluster) %>% top_n(1, abs(avg_log2FC))
  

p2 <- FeaturePlot(object = all.samples.integrated, 
                  shape.by = 'spp',
                  split.by = 'spp',
                  label = T, pt.size = 0.6, label.size = 3, 
                  features = unique(matched.DEGs.top$gene),
                  cols = c("lightgrey", "red"), reduction = "pca") 

plot(p2)
```

```{r matched_expression_down2, eval = T, echo = F, message=FALSE, warning=FALSE,fig.width=8, fig.height=8}
p3 <- DotPlot(all.samples.integrated.sub, features = unique(matched.DEGs.top$gene), 
              cols = c("blue", "red", "green", "yellow"),
              dot.scale = 8) + RotatedAxis()

plot(p3)
```

```{r matched_expression_down3, eval = T, echo = F, message=FALSE, warning=FALSE,fig.width=12, fig.height=8}
VlnPlot(object = all.samples.integrated.sub, features = unique(matched.DEGs.top$gene))
```

### Clustering DEGs

1. Hierarchical clustering was performed per phase and expression heatmaps were generated

```{r Matched_Heatmap, eval = T, echo = F, message=FALSE, warning=FALSE, fig.width=6, fig.height=6}
## Matched
## Get the average expression per timepoint
Idents(all.samples.integrated) <- 'phase.cond'
DefaultAssay(all.samples.integrated) <- 'RNA'
matched.avg.expr <- as.data.frame(log1p(AverageExpression(all.samples.integrated, verbose = FALSE)$RNA))
matched.avg.expr$gene <- rownames(matched.avg.expr)
matched.avg.expr$GeneID <- gsub('-', '_', matched.avg.expr$gene)

matched.up.reg <- matched.DEGs.sig$GeneID[matched.DEGs.sig$avg_log2FC > 0]
matched.down.reg <- matched.DEGs.sig$GeneID[matched.DEGs.sig$avg_log2FC < 0]

matched.avg.expr.up <- matched.avg.expr[matched.avg.expr$GeneID %in% matched.up.reg, ]
matched.avg.expr.down <- matched.avg.expr[matched.avg.expr$GeneID %in% matched.down.reg, ]


matched.mat_up <- matched.avg.expr.up[,1:(ncol(matched.avg.expr.up) - 2)] %>% dplyr::select(!contains('Y'))
matched.mat_down <- matched.avg.expr.down[,1:(ncol(matched.avg.expr.down) - 2)] %>% dplyr::select(!contains('Y'))


L1 <- makeMatchedHeatMap(matched.mat_up, prod.desc, 7)
L2 <- makeMatchedHeatMap(matched.mat_down, prod.desc, 7)


draw(L1$ht_list)
draw(L2$ht_list)
```